Predicting the PCG PSeudo-Random Number Generator In Practice

### 2020-05-27 article/main.tex
refonte section 3

\section{Algorithm of Frieze \textit{et al}}

\todo{à fusionnier avec la section suivante!}
Pour casser PCG, nous avons utilisé la méthode présentée par Alan M. Frieze, Johan Hastad, Ravi Kannan, Jeffray C. Lagarias et Adi Shamir \cite{Frieze}. Plaçons nous dans le cas simple du Lehmer's generator \todo{FIXME}. It's a Linear Congruential Generator, a known multiplier $A$, a known modulus $2^128$ and no increment. The outputs $X$ are the 64 upper-bits of $S$. For $n \in \mathbb{N}$, the $n$ firsts internal states form a lattice we will name $\mathcal{L}(A,2^{128},n)$. 
\todo{rajouter ref Lehmer's generator}

This lattice is generated by the matrix :
\begin{equation}
genMatrix =
\begin{pmatrix} 
1 & A & A^2 & \dots & A^{n- 1}\\
0 & M & 0 & \dots & 0\\
0 & 0 & M & \dots & 0\\
\vdots & & & \ddots & \vdots\\
0 & 0 & \dots & 0 & M
\end{pmatrix}
\end{equation}

The idea of this method is to consider a known vector $T$, close to $S$ the internal state we are looking for, and to search for the closest vector in $\mathcal{L}(A,2^{128},n)$. In the case of Lehmer's generator, we use $X * 2^{64}$ as an approximation of $S$.

\todo{preuve que c'est bien le plus proche ici ?}

Pour améliorer les performances de nos algorithmes les plus critiques, nous n'avons dans certains cas pas utilisé d'oracle CVP parfait, mais la méthode du babai-rounding.

\todo{description babai rounding}
\todo{et preuve ?}

@@

Given a bit-size $k$, a fixed multiplier $a$ and an increment $c$, define the
sequence $u_0 = x, u_{i+1} = a u_i + c \bmod 2^k$. The sequence $u$ form the
successive internal states of a linear congruential generator (LCG). When the
increment is zero, the sequence forms a geometric progression with common ratio
$a$, and is sometimes referred to as the output of a multiplicative congruential
generator (MCG).

Let $\LCG_{n,k}(x, c)$ denote the vector
$(u_0, \dots, u_{n-1}) \in \mathbb{Z}_{2^k}$ (we will omit $n$ and $k$ when they
are obvious from the context).  It is easy to check that:
\begin{align}
  \label{eq:lcg-additive}
  \LCG_{n,k}(x + y, c + d) &= \LCG_{n,k}(x, c) + \LCG_{n,k}(y, d),  \\
  \label{eq:lcg-scalar}
  \LCG_{n,k}(\lambda x, \lambda c) &= \lambda \LCG_{n,k}(x, c).   
\end{align}

\todo{à fusionnier avec la section précédente!}
\subsection{The lattice}

One of the key point of the algorithm is, given a modulus \(M\) and a given common ration \(\alpha\), to find a modular geometric progression knowing only the most significant bits of its \(n\) first terms.

We call \(L(\alpha,M,n)\) the set of n-tuples \(U = (u_0,\dots u_{n-1})\) satisfying :

\[u_{i} = \alpha u_{i-1} \mod M \] for all \(i\) in \(\{1,\dots,n-1\}\).

We easily notice that, if \(U\) and \(V\) are in \(L(\alpha,M,n)\), so are \(-U\) and \(U+V\), hence \(L(\alpha,M,n)\) is a lattice.

The lattice \(L(\alpha,M,n)\) is spawned by the lines of the following \(n\times n\) matrix :

\[\begin{pmatrix}
1&\alpha&\alpha^2&\dots&\alpha^{n-1}\\
0&M&0&\dots&0\\
0&0&M&\dots&0\\
\dots&\dots&\dots&\dots&\dots\\
0&0&0&\dots&M\\
\end{pmatrix}\]

@@

We know that \(U\) is in the lattice, \(T\) is usually not and \(|T-U|\) is small. Then we can hope for \(U\) to be the closet vector of \(T\) which is in the lattice : the vector \(U\) may be the solution of \(CVP(L(\alpha,M,n),T)\).

What can we say about \(|CVP(L(\alpha,M,n),T)-U|\) ?

\begin{align*}
|CVP(L(\alpha,M,n),T)-U| &= |CVP(L(\alpha,M,n),T)-T+T-U|\\
& \leqslant |CVP(L(\alpha,M,n),T)-T|+|T-U|\\
\end{align*}

But, as the vector \(U\) is in the lattice and by definition of the CVP, \(|CVP(L(\alpha,M,n),T)-T| \leqslant |T-U| \), hence 
\[|CVP(L(\alpha,M,n),T)-U|\leqslant 2|T-U|\]

If we can prove that the right side of the inequality is smaller than the SVP of the lattice, then wa have proved that \(CVP(L(\alpha,M,n),T)\) is indeed the vector we were searching for.

@@

As using an exact CVP solver is costly, we'll try a cheaper but less effective way to find the closest vector.

We compute \(G\) the LLL-reduction of the generating matrix of \(L(\alpha,M,n)\). We denote by  \(\rVert G \lVert\) the induced matrix norm :
\[\rVert G \lVert =  \max_{x \in \mathbb{R}}\frac{\lVert Ax \rVert_2}{\lVert x \rVert_2}\]
In the case of the \(\lVert\rVert_2\) norm, \(\rVert G \lVert = \sigma(G)\), where \(\sigma(G)\) is the largest singular value of \(G\).

@@

If we can prove that the right side of the inequality is smaller than the SVP of G, then wa have proved that \(rG\) is indeed the vector we were searching for.

We see that  the right side of the inequality is often larger than in the first case. We will need to use larger parameters to find the vector we are looking for. But as the Babai rounding is largely cheaper than using an exact CVP solver, there will be cases where this method is far more practical. \todo{je trouve la formulation un peu moche}

@@

\begin{proof} \todo{FIX ME!}
The vector \(S'\) has been constructed such that it is in \(L(A,2^{128},n)\) and its \(\ell\) lower bits are zero. As \(S'\) is in \(L(A,2^{128},n)\), \(S' \mod [2^{64+\ell}]\) is in \(L(A,2^{64+\ell},n)\)

We call \(U\) the truncated vector : \(S'[\ell,64+\ell]\). As the \(\ell\) lower bits of \(S'\) are zero, we have \(S' \equiv U \times 2^\ell\ \mod [2^{64+\ell}]\). Hence \(U\) is in \(L(A,2^{64},n)\).

We consider the case were we successfully guessed the 6 upper bits of each \(S_i\) and the \(\ell\) lower bits of \(S_0\). Then we know the \(6+\ell\) upper bits of \(U\).

\todo{maybe dessin}

We consider the vector \(T'\) computed in step (numerodelastepenquestion). To recover \(U\), we use the Babai's rounding explain in section (insererlasection) on \(T'\) and \(L(A,2^{64},n)\).

As \(U\) and \(T'\) have exactly the same \(6+\ell\) upper bits  know that \(\lVert U-T \rVert \leqslant \sqrt{n2^{2(58-l)}}\).

\[\sqrt{n2^{2(58-\ell)}} \times (1 +\lVert G^{-1} \rVert  \times \lVert G\rVert) \leqslant \lVert SVP(L(A,2^{64},n))\rVert \]

\begin{itemize}
	\item For \(n=3\)

	\(G = \begin{pmatrix}
	-1241281756092&3827459685972&-728312298332\\
	-5001120657083&-2117155768935&5479732607037\\
	8655886039732&3303731088004&6319848582548\\
	\end{pmatrix}\)

	\(\lVert G \rVert \times \lVert G^{-1} \rVert = 2.869065991049632\)

	\(\lVert SVP(G) \rVert = 4.08909120094543e{^12}\)

	Hence, the smallest l such that the inequality is satisfied is \(\ell = 19\) %17 avec borne CVP

	\item For \(n=4\)

	\(G = \begin{pmatrix}
	-186304953996472&-126056243766680&7937589136904&93078431381544\\
	-216211368070119&99587582169277&-214303762177807&-1707551230219\\
	110964501361298&-5646098666150&-268280113597118&149382085707466\\
	131252974561432&-233919070109448&-98716819647784&-134620659538888\\
	\end{pmatrix}\)

	\(\lVert G \rVert \times \lVert G^{-1} \rVert = 2.060717188416003\)

	\(\lVert SVP(G) \rVert = 2.43569932844650e^{14}\)

	Hence, the smallest l such that the inequality is satisfied is \(\ell = 13\) %11 avec borne CVP

	\item For \(n=5\)

	\(\lVert G \rVert \times \lVert G^{-1} \rVert = 3.7692142154392747\)

	\(\lVert SVP(G) \rVert = 1.71669732658884e^{15}\)

	Hence, the smallest l such that the inequality is satisfied is \(\ell = 11\)  %9 avec borne CVP

	\item For \(n=6\)

	\(\lVert G \rVert \times \lVert G^{-1} \rVert = 2.6925925483728537\)

	\(\lVert SVP(G) \rVert =1.02550589786192e^{16}\)

	Hence, the smallest l such that the inequality is satisfied is \(\ell = 8\) %7 avec borne CVP
\end{itemize}
\end{proof}

>>>>

@@

Given an integer \(k\), a fixed multiplier \(a\) and an increment \(c\), define the
sequence \[u_0 = x, u_{i+1} = a u_i + c \bmod 2^k\] The sequence \(u\) form the
successive internal states of a linear congruential generator (LCG).

Let $\LCG_{n,k}(x, c)$ denote the vector $(u_0, \dots, u_{n-1}) \in \mathbb{Z}_{2^k}$ (we will omit $n$ and $k$ when they are obvious from the context).  It is easy to check that:
\begin{align}
\label{eq:lcg-additive}
\LCG_{n,k}(x + y, c + d) &= \LCG_{n,k}(x, c) + \LCG_{n,k}(y, d),  \\
\label{eq:lcg-scalar}
\LCG_{n,k}(\lambda x, \lambda c) &= \lambda \LCG_{n,k}(x, c).   
\end{align}
Hence those sets of vector have a lot of structure we can use.

The LCG have been largely studied. This is why we know techniques to recover internal states knowing only upper or lower bits of the output, see \cite{Boyar1989}. A particular LCG is the Lehmer's generator \cite{Lehmer} \todo{je sais pas écire la référence dans bibtex}, where the multiplier \(a\) is known and the increment is zero.

In our algorithm to predict the PCG, we need to reconstruct several truncated Lehmer's generator. For that, we will present a classical lattice-based reconstruction for a truncated Lehmer's generator, as presented in \cite{Frieze}. \todo{C'est pas exactement celle là, on dit quoi du coup ?}

\subsection{The lattice associated to a Lehmer's generator}

We fix a multiplier \(a\), integers \(k\), \(n\) and we consider \(L(n,k)\) the set of \(\LCG_{n,k}(x, 0)\), for \(x\) in \(\mathbb{Z}_{2^k}\). Then

\begin{itemize}
	\item By \ref{eq:lcg-additive}, the vector \(\LCG(x_1,0) + \LCG(x_2,0) = \LCG(x_1+x_2,0)\), hence it is still in \(L(n,k)\).
	\item By \ref{eq:lcg-scalar}, the vector \((-1)\times \LCG(x,0) = \LCG(-x,0)\) hence it is still in \(L(n,k)\).
\end{itemize}

Hence \(L(n,k)\) is a lattice. We easily notice that \(L(n,k)\) is spawned by the lines of the following \(n\times n\) matrix :

\[\begin{pmatrix}
1&a&a^2&\dots&a^{n-1}\\
0&2^k&0&\dots&0\\
0&0&2^k&\dots&0\\
\dots&\dots&\dots&\dots&\dots\\
0&0&0&\dots&2^k\\
\end{pmatrix}\]

Once again, we will omit $n$ and $k$ when they are obvious from the context.

@@

We know that \(U\) is in the lattice, \(T\) is usually not and \(|T-U|\) is small. Then we can hope for \(U\) to be the closet vector of \(T\) which is in the lattice : the vector \(U\) may be the solution of \(CVP(L,T)\).

What can we say about \(|CVP(L,T)-U|\) ?

\begin{align*}
|CVP(L,T)-U| &= |CVP(L,T)-T+T-U|\\
& \leqslant |CVP(L,T)-T|+|T-U|\\
\end{align*}

But, as the vector \(U\) is in the lattice and by definition of the CVP, \(|CVP(L,T)-T| \leqslant |T-U| \), hence 
\[|CVP(L,T)-U|\leqslant 2|T-U|\]

If we can prove that the right side of the inequality is smaller than first minimum of the lattice \(\lambda_1(L)\), then we have proved that \(CVP(L,T)\) is indeed the vector we were searching for.

In this article, we will only use the Babai's rounding when \(k = 128-\ell\) and when we know the \(6\) upper bit of our Lehmer's generator. Hence \(\lVert T-U \rVert \leqslant \sqrt{n}2^{122-\ell} \). So, if we fix \(\ell\) we can search for the minimum \(n\) such that \[2\sqrt{n}2^{122-\ell} \leqslant \lambda_1(L(128-\ell,n))\].

\begin{tabular}{|c|c|}
	\hline
	\(\ell\)  & minimum \(n\) \\
	\hline
	11 & 39 \\
	12 & 38 \\
	13 & 38 \\
	\hline
\end{tabular}

@@

As using an exact CVP solver is costly, we'll try a cheaper but less precise way to find the closest vector.

We compute \(G(n,k)\) the LLL-reduction of the generating matrix of \(L(n,k)\). We denote by  \(\rVert G \lVert\) the induced matrix norm :
\[\rVert G \lVert =  \max_{x \in \mathbb{R}}\frac{\lVert Gx \rVert_2}{\lVert x \rVert_2}\]
In the case of the \(\lVert\rVert_2\) norm, \(\rVert G \lVert = \sigma(G)\), where \(\sigma(G)\) is the largest singular value of \(G\).

@@

If we can prove that the right side of the inequality is smaller than the minimum of the lattice \(\lambda_1(L)\), then we have proved that \(rG\) is indeed the vector we were searching for.

In this article, we will only use the Babai's rounding when \(k = 64\) and when we know the \(6+\ell\) upper bit of our Lehmer's generator. Hence \(\lVert T-U \rVert \leqslant \sqrt{n}2^{58-\ell} \). So, if we fix \(n\) we can search for the minimum \(\ell\) such that \[(1+\lVert G(n,64) \rVert \times \lVert G(n,64)^{-1} \rVert)\sqrt{n}2^{58-\ell} \leqslant \lambda_1(L(n,64))\]

\bigskip

\begin{tabular}{|c|c|c|c|}
	\hline
	n & \(\lVert G(n,64) \rVert \times \lVert G(n,64)^{-1} \rVert\) & \( \lambda_1(L(n,64)) \) & minimum \(\ell\) \\
	\hline
	3 & 2.869065991049632 & \(4.08909120094543e^{12}\) & 19 \\
	4 & 2.060717188416003 & \(2.43569932844650e^{14}\) & 13 \\
	5 & 3.7692142154392747 & \(1.71669732658884e^{15}\) & 11 \\
	6 & 2.6925925483728537 & \(1.02550589786192e^{16}\) & 8 \\
	\hline
\end{tabular}

@@

\todo{avec la nouvelle partie 3 et le blabla au dessus de l'algo, y'a plus vraiment besoin de preuve}
